---
title: "NetPerturb Report"
subtitle: "Perturbation scores across targets and cell types"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    number-sections: true
    page-layout: full
    theme: flatly
params:
  perbscore_file: "perbscore_all_targets.txt"
  umap_file: "umap.png"
  network: ""
  score_quantile: 0.75
  sctknk_file: ""
  sctknk_top_genes: 25
  gsea_file: ""
  gsea_top_terms: 20
  target_qc_file: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(scales)

# plotly draws the knockout networks. A container built before that section
# existed still has to render the rest of the report, so its absence is a
# missing figure rather than an error.
has_plotly <- requireNamespace("plotly", quietly = TRUE)

perb_score <- read.delim(params$perbscore_file, header = TRUE, sep = "\t")

n_cell_types <- n_distinct(perb_score$cell_type)
n_targets    <- n_distinct(perb_score$target)
n_bindings   <- n_distinct(perb_score$binding)

# Pool every score from every target on a log10 scale; non-positive scores
# cannot be logged and are carried as NA.
score_data <- perb_score %>%
  mutate(log_score = ifelse(perb_score > 0, log10(perb_score), NA_real_))

has_scores <- any(is.finite(score_data$log_score))

# Pooled scores are bimodal: a low mode of near-zero scores from genes the
# network barely leans on, and a high mode that carries the signal. Imputing
# the low mode up to a bound only stacked it against that bound and left it in
# the picture, so it is cut instead. Everything below the score_quantile
# quantile of the pooled log10 scores is dropped and the high mode is plotted
# on its own; the default 0.75 cuts at q3.
score_quantile <- suppressWarnings(as.numeric(params$score_quantile))
if (!isTRUE(is.finite(score_quantile)) || score_quantile < 0 || score_quantile >= 1) {
  score_quantile <- 0.75
}

score_cut <- if (has_scores) {
  unname(quantile(score_data$log_score, score_quantile, na.rm = TRUE))
} else {
  NA_real_
}

score_data <- score_data %>%
  mutate(log_score_kept = ifelse(
    is.finite(log_score) & log_score >= score_cut, log_score, NA_real_
  ))

n_scored  <- sum(is.finite(score_data$log_score))
n_kept    <- sum(is.finite(score_data$log_score_kept))
n_below   <- n_scored - n_kept
n_dropped <- sum(!is.finite(score_data$log_score))

# The strongest scores in the run. "Overall" is the whole pooled set, the same
# population the retention cut is taken from, so this threshold does not move
# with score_quantile. Change top_quantile alone to mark a different fraction.
top_quantile <- 0.975
top_label    <- paste0("Top ", signif((1 - top_quantile) * 100, 3), "%")

top_cut <- if (has_scores) {
  unname(quantile(score_data$log_score, top_quantile, na.rm = TRUE))
} else {
  NA_real_
}

score_data <- score_data %>%
  mutate(is_top = is.finite(log_score) & log_score >= top_cut)

n_top <- sum(score_data$is_top)

# The network inference method the run was launched with (--network). The
# report also has to render when it is not supplied, e.g. when the qmd is
# knitted by hand against a table someone already has.
network_key   <- if (is.null(params$network)) "" else trimws(as.character(params$network))
network_names <- c(
  genie3  = "GENIE3",
  scrank  = "scRank",
  hdwgcna = "hdWGCNA"
)
network_label <- unname(network_names[network_key])
if (!nzchar(network_key)) {
  network_label <- "an unreported method"
} else if (is.na(network_label)) {
  network_label <- network_key
}
network_value <- if (nzchar(network_key)) network_key else "not supplied"

# Target genes DOWNSAMPLE took out of the run before any network was built:
# absent from the object, or with no counts in the cells it kept. Missing or
# empty when the qmd is rendered by hand, or by a run from before that check,
# which reads the same as a run where every gene passed.
target_qc_path <- params$target_qc_file
target_qc <- if (!is.null(target_qc_path) && nzchar(target_qc_path) &&
                 file.exists(target_qc_path) && file.size(target_qc_path) > 0) {
  read.delim(target_qc_path, header = TRUE, sep = "\t",
             colClasses = "character", check.names = FALSE)
} else {
  data.frame(target = character(), gene = character(), reason = character(),
             action = character())
}
has_target_qc <- nrow(target_qc) > 0

# The UMAP is a QC figure produced upstream by DOWNSAMPLE. A run started before
# that step existed, or one whose plot could not be drawn, still has to render.
umap_path <- params$umap_file
has_umap  <- !is.null(umap_path) && nzchar(umap_path) && file.exists(umap_path)
if (has_umap) umap_path <- normalizePath(umap_path)

# scTenifoldKnk's differentially-regulated (DR) gene table, from a --sctknk
# run. A header-only sentinel file (no rows) stands in when it was not
# requested for this run at all, same shape as a real run that just found
# nothing, so both render the same "nothing to show" message.
# The named-but-empty frame matters: everything below filters on p.adj, so a
# bare data.frame() with no columns would fail with "object 'p.adj' not found"
# rather than fall through to the "not run" message. Same reasoning, and the
# same 8 columns, as empty_dr in bin/sctenifoldknk_ko.R. The size check is for
# MERGE_SCTENIFOLDKNK's stub, which touches a zero-byte file that read.delim
# refuses with "no lines available in input".
sctknk_empty <- data.frame(
  cell_type = character(), target = character(), gene = character(),
  distance = numeric(), Z = numeric(), FC = numeric(),
  p.value = numeric(), p.adj = numeric(), stringsAsFactors = FALSE
)

sctknk_path <- params$sctknk_file
sctknk_dr <- if (!is.null(sctknk_path) && nzchar(sctknk_path) &&
                  file.exists(sctknk_path) && file.size(sctknk_path) > 0) {
  read.delim(sctknk_path, header = TRUE, sep = "\t")
} else {
  sctknk_empty
}

has_sctknk <- nrow(sctknk_dr) > 0

# How many genes the network figure draws around each target. The table below
# it is never capped.
sctknk_top_genes <- suppressWarnings(as.integer(params$sctknk_top_genes))
if (!isTRUE(is.finite(sctknk_top_genes)) || !isTRUE(sctknk_top_genes >= 1)) {
  sctknk_top_genes <- 25L
}

# dRegulation() reports FC as a gene's squared manifold-alignment distance over
# the mean squared distance of the run, so it is a ratio to the average rather
# than a differential-expression fold change, and it carries no direction: a
# gene that moved a long way scores high whichever way it moved. log2 of it is
# what the figure sizes nodes by and what the table reports.
if (has_sctknk) {
  sctknk_dr <- sctknk_dr %>%
    mutate(log2FC = ifelse(is.finite(FC) & FC > 0, log2(FC), NA_real_))
} else {
  # No rows, but the 8 columns are there either way, so only log2FC has to be
  # added to keep the select() in the table below working.
  sctknk_dr$log2FC <- numeric()
}

# p.value is pchisq(FC, df = 1) and p.adj is BH over that, so the FDR cut is a
# monotone function of FC: filtering here is the same as taking everything
# above a log2FC threshold. Both sections below start from this frame.
sctknk_sig <- sctknk_dr %>%
  filter(!is.na(p.adj), p.adj < 0.05) %>%
  arrange(target, cell_type, p.value)

has_sctknk_sig <- nrow(sctknk_sig) > 0
n_sctknk_sig   <- nrow(sctknk_sig)

# Counted over the significant rows, not over everything knocked out: a target
# that came back with nothing would otherwise be counted in a sentence that
# says how far the hits reach.
n_sctknk_targets    <- n_distinct(sctknk_sig$target)
n_sctknk_cell_types <- n_distinct(sctknk_sig$cell_type)

# Ring nodes for the figure. SCTENIFOLDKNK_KO already drops the knocked-out
# gene(s) from the DR table -- they always move furthest by construction -- so
# this filter only matters for a table from a run before that; it would draw
# them as spokes back to themselves. A ';'-joined target is a joint knockout,
# so all of its genes are dropped. The ring is then capped, since a target can clear FDR on
# hundreds of genes and every view is embedded in the HTML up front.
is_ko_gene <- function(gene, target) {
  mapply(function(g, t) g %in% strsplit(t, ";", fixed = TRUE)[[1]], gene, target)
}

sctknk_net <- if (has_sctknk_sig) {
  sctknk_sig %>%
    filter(!is_ko_gene(gene, target), is.finite(log2FC)) %>%
    group_by(target, cell_type) %>%
    slice_max(log2FC, n = sctknk_top_genes, with_ties = FALSE) %>%
    ungroup()
} else {
  sctknk_sig
}

has_sctknk_net <- nrow(sctknk_net) > 0

# Counted off sctknk_net rather than sctknk_sig, because that is the frame the
# dropdown is built from: in a table from before the knocked-out genes were
# dropped upstream, a combination whose only significant gene was the
# knocked-out gene itself has a row in the table but no menu entry.
n_sctknk_views <- if (has_sctknk_net) {
  nrow(distinct(sctknk_net, target, cell_type))
} else {
  0
}

# The enrichment run over the table above, from --gsea_gmt. Same sentinel
# arrangement and the same reasons: assets/NO_GSEA_TABLE is a header-only file
# standing in for "no gene sets were given", the named-but-empty frame keeps the
# filters below from failing on a missing column, and the size check is for the
# stub's zero-byte file.
gsea_empty <- data.frame(
  cell_type = character(), target = character(), pathway = character(),
  pval = numeric(), padj = numeric(), log2err = numeric(),
  ES = numeric(), NES = numeric(), size = integer(),
  leadingEdge = character(), stringsAsFactors = FALSE
)

gsea_path <- params$gsea_file

# Unlike the knockout table, this one has two ways of being empty that want
# different sentences: no --gsea_gmt was given at all, or one was and nothing
# came back -- which is what a GMT for the wrong species looks like from here.
# The sentinel is staged under its own name, so the file name tells them apart.
gsea_requested <- !is.null(gsea_path) && nzchar(gsea_path) &&
  basename(gsea_path) != "NO_GSEA_TABLE"

gsea <- if (!is.null(gsea_path) && nzchar(gsea_path) &&
              file.exists(gsea_path) && file.size(gsea_path) > 0) {
  read.delim(gsea_path, header = TRUE, sep = "\t")
} else {
  gsea_empty
}

has_gsea <- nrow(gsea) > 0

gsea_top_terms <- suppressWarnings(as.integer(params$gsea_top_terms))
if (!isTRUE(is.finite(gsea_top_terms)) || !isTRUE(gsea_top_terms >= 1)) {
  gsea_top_terms <- 20L
}

# 0.25, not the 0.05 the DR table above is cut at. That is GSEA's own
# convention -- the Broad's guidance for the method sets significance at FDR
# q < 0.25 -- and it exists because a gene set test is not a gene test: the
# sets in a collection overlap heavily, so the BH correction over them is
# conservative, and the question here is which processes to follow up rather
# than which single genes to believe.
gsea_fdr <- 0.25

gsea_sig <- gsea %>%
  filter(!is.na(padj), padj < gsea_fdr) %>%
  mutate(
    # The leading edge is stored ';'-joined; its length is what the figure
    # sizes dots by and the only part of it the table keeps. as.character()
    # rather than the column as read: read.delim types a column of a
    # header-only table -- which is exactly what assets/NO_GSEA_TABLE is --
    # as logical, and strsplit() refuses a non-character argument, so without
    # it a run whose enrichment found nothing fails to render at all.
    leading_edge_n = lengths(strsplit(as.character(leadingEdge), ";", fixed = TRUE)),
    # The figure's colour. -log(p.adj + 1), not -log10(p.adj): the pseudocount
    # bounds it -- p.adj in (0, 1] puts this in [-log(2), 0), rising towards 0
    # as a set gets more significant -- so fgsea's 1e-50 p-value floor cannot
    # stretch the scale any more, and no cap is needed. The trade is at the
    # significant end: log is close to linear near 1, so everything below
    # p.adj of about 0.01 lands within a hair of 0 and takes the same colour.
    # The exact p.adj is a column of the table below.
    neg_log_padj = -log(as.numeric(padj) + 1)
  ) %>%
  arrange(target, cell_type, padj)

has_gsea_sig      <- nrow(gsea_sig) > 0
n_gsea_sets       <- nrow(gsea_sig)
n_gsea_targets    <- n_distinct(gsea_sig$target)
n_gsea_cell_types <- n_distinct(gsea_sig$cell_type)

# Set names come out of the GMT in MSigDB's shape -- GOBP_REGULATION_OF_... --
# which is unreadable at axis size, and long enough to be unreadable in a table
# column too. Underscores become spaces for the axis label only; the hover
# keeps the name exactly as it was given, since that is what anyone would
# search the collection for, and the table keeps it in the cell behind a
# truncating renderer so its filter still matches the part it does not show.
#
# One budget for both, so a name is not cut at two different lengths in one
# section. The table's renderer is JavaScript and carries the number itself.
set_label_max <- 50

pretty_set <- function(x) {
  label <- gsub("_", " ", x)
  ifelse(nchar(label) > set_label_max,
         paste0(substr(label, 1, set_label_max - 1), "\u2026"), label)
}

# Figure rows: the strongest sets per combination. Capped for the same reason
# the knockout ring is -- every view is embedded in the HTML up front -- while
# the table below stays uncapped. Ranked on |NES| rather than NES: the ranking
# behind these scores is two-tailed, so a set at the quiet end of it is as much
# a result as one at the moved end, and slicing on NES alone would keep only
# the moved end and silently drop the other half of the figure.
gsea_fig <- if (has_gsea_sig) {
  gsea_sig %>%
    filter(is.finite(NES), is.finite(neg_log_padj)) %>%
    group_by(target, cell_type) %>%
    slice_max(abs(NES), n = gsea_top_terms, with_ties = FALSE) %>%
    ungroup()
} else {
  gsea_sig
}

has_gsea_fig <- nrow(gsea_fig) > 0

n_gsea_views <- if (has_gsea_fig) {
  nrow(distinct(gsea_fig, target, cell_type))
} else {
  0
}
```

## Overview

This report summarizes `r nrow(perb_score)` perturbation scores from
`r params$perbscore_file`, covering `r n_targets` target(s) across
`r n_cell_types` cell type(s), scored under `r n_bindings` binding
mode(s) (`r paste(sort(unique(perb_score$binding)), collapse = ", ")`).

The gene regulatory networks these scores were computed from were inferred with
**`r network_label`** (`--network`: `r network_value`). Every score below is
relative to that network type, so scores from a run using a different inference
method are not directly comparable.

### Target QC

```{r target-qc-note, results="asis"}
if (has_target_qc) {
  cat("These genes were not analyzed due to QC checking: ",
      paste0("**", unique(target_qc$gene), "**", collapse = ", "),
      ". A gene is excluded when it is absent from the expression profile or ",
      "has zero counts in every cell retained after downsampling, since it can ",
      "have no edges in any network. A combined target loses only the failing ",
      "gene and is analysed with the rest.\n", sep = "")
} else {
  cat("Every requested target gene passed QC checking.\n")
}
```

```{r target-qc-table, eval=has_target_qc}
datatable(
  target_qc,
  rownames = FALSE,
  colnames = c("Requested target", "Gene", "Reason", "Action"),
  options = list(dom = "t", pageLength = nrow(target_qc))
)
```

## Perturbation score table

Use the column filters below to query by cell type, target, or binding.

```{r table}
perb_score %>%
  arrange(target, cell_type) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    options = list(pageLength = 15, order = list(list(1, "asc"))),
    caption = "All target x cell type perturbation scores."
  ) %>%
  formatSignif(columns = "perb_score", digits = 4)
```

## Perturbation score distribution

All `r n_scored` log10 scores from every target and cell type pooled into one
distribution. Pooled scores are typically bimodal: a low mode of near-zero
values from genes the network barely leans on, and a high mode carrying the
signal. Only the high mode is kept, cut at the `r score_quantile` quantile
(red line, `r if (has_scores) signif(score_cut, 4) else NA`). That retains
`r n_kept` score(s) and drops `r n_below` below the cut; a further `r n_dropped`
non-positive score(s) cannot be logged and are not shown either.

The left panel is every score, so the cut can be checked against where the two
modes actually separate — move it with `--score_quantile` if the red line does
not fall in the valley between them. The right panel is what the heatmap below
works with, on its own axis.

The green dashed line in both panels is the `r top_label` threshold
(`r if (has_scores) signif(top_cut, 4) else NA`), the `r n_top` strongest
score(s) in the run. Those are the scores the heatmap asterisks mark.

```{r distribution, fig.width=10, fig.height=4, eval=has_scores}
dist_data <- score_data %>%
  select(log_score, log_score_kept) %>%
  pivot_longer(everything(), names_to = "stage", values_to = "value") %>%
  filter(is.finite(value)) %>%
  mutate(stage = factor(
    stage,
    levels = c("log_score", "log_score_kept"),
    labels = c("All scores", "Retained high mode")
  ))

# The retention cut belongs only to the panel it acts on; the top threshold is
# drawn in both, so the strongest scores can be placed against the whole set
# and against the mode the report keeps.
stages <- levels(dist_data$stage)

ref_lines <- rbind(
  data.frame(
    stage = factor("All scores", levels = stages),
    xintercept = score_cut,
    line = "Quantile cut"
  ),
  data.frame(
    stage = factor(stages, levels = stages),
    xintercept = top_cut,
    line = top_label
  )
)

# Red and green are a poor pair for the commonest colour blindness, so the two
# lines differ in dash as well as hue and the legend names both.
ref_colours   <- setNames(c("#e34948", "#008300"), c("Quantile cut", top_label))
ref_linetypes <- setNames(c("solid", "dashed"), c("Quantile cut", top_label))

# A density curve is what makes the two modes readable, but it needs a few
# points per panel to mean anything.
show_density <- min(table(dist_data$stage)) >= 3

ggplot(dist_data, aes(x = value)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 20, fill = "steelblue", color = "white"
  ) +
  { if (show_density) geom_density(color = "grey25", linewidth = 0.6) } +
  geom_vline(
    data = ref_lines,
    aes(xintercept = xintercept, colour = line, linetype = line),
    linewidth = 0.7
  ) +
  scale_colour_manual(name = NULL, values = ref_colours) +
  scale_linetype_manual(name = NULL, values = ref_linetypes) +
  facet_wrap(~stage, scales = "free") +
  labs(
    title = "Distribution of perturbation scores across all targets",
    x = "log10(perturbation score)",
    y = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r distribution-empty, eval=!has_scores, results="asis"}
cat("No positive perturbation score is available, so no distribution is shown.")
```

## Perturbation score heatmap

Scores are log10-transformed and cut as above, so the colour scale spans the
high mode only. Tiles below the cut, and those whose score is non-positive, are
shown as missing (gray). An asterisk marks a score in the `r top_label` of every
score in the run — `r n_top` tile(s), at or above
`r if (has_scores) signif(top_cut, 4) else NA` on the log10 scale.

```{r heatmap, fig.width=10, fig.height=max(4, 0.4 * n_cell_types)}
ggplot(score_data, aes(x = target, y = cell_type, fill = log_score_kept)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(
    data = filter(score_data, is_top),
    aes(label = "*"), colour = "#0b0b0b", size = 6, vjust = 0.75
  ) +
  scale_fill_viridis_c(name = "log10(perb_score)", na.value = "gray80") +
  labs(x = "Target", y = "Cell type") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  { if (n_bindings > 1) facet_wrap(~binding) }
```

## Knockout networks

Each knockout and the genes it moved. scTenifoldKnk zeroes the target's
outgoing edges, realigns the knocked-out network against the wild-type
one, and scores every gene on how far it travelled between the two; the genes
drawn here are the ones clearing FDR < 0.05. Use the menu to switch between the
`r n_sctknk_views` target and cell type combinations.

The knocked-out gene sits at the centre and the genes it moved sit around it. A
combined target such as `A;B` is a single joint knockout, with both genes' edges
zeroed together, so all of its genes sit at the centre, each linked to every gene
the knockout moved. A node's size follows its `log2FC`, scaled within the combination on display, so
sizes rank genes against each other rather than across cell types. The ring is
capped at the `r sctknk_top_genes` strongest genes per combination; the table in
the next section is not capped. Hover any node for its exact values.

`log2FC` here is the log2 ratio of a gene's squared alignment distance to the
mean squared distance of its run, so it says how far a gene moved and not which
way it went. It carries no direction, which is why no edge on this figure has a
sign to read off: every gene shown simply moved further than its run's average.

```{r sctknk-network, eval=has_sctknk_net && has_plotly}
# log2FC has no sign and is already carried by node size, so nothing on this
# figure needs to carry meaning through colour and the palette stays neutral.
edge_colour <- "#8f8e8a"
node_fill   <- "#e8e7e3"
node_line   <- "#52514e"
target_fill <- "#3f3e3b"
surface     <- "#fcfcfb"

# The subnetwork of one knockout is a star, so it needs no force layout: the
# knocked-out gene goes in the middle and the genes it moved onto a ring.
star_positions <- function(edges) {
  centre <- sort(unique(edges$target_gene))
  ring   <- setdiff(unique(edges$partner), centre)

  centre_pos <- if (length(centre) == 1) {
    data.frame(gene = centre, x = 0, y = 0, label_at = "bottom center")
  } else {
    angle <- (seq_along(centre) - 1) * 2 * pi / length(centre) + pi / 2
    data.frame(gene = centre, x = 0.22 * cos(angle), y = 0.22 * sin(angle),
               label_at = "bottom center")
  }

  ring_pos <- if (length(ring) > 0) {
    angle <- (seq_along(ring) - 1) * 2 * pi / length(ring) + pi / 2
    # A label pushed straight out from the centre clears both its own node and
    # the edge running into it, which "top center" on every node does not.
    label_at <- ifelse(
      cos(angle) > 0.5, "middle right",
      ifelse(cos(angle) < -0.5, "middle left",
             ifelse(sin(angle) > 0, "top center", "bottom center"))
    )
    data.frame(gene = ring, x = cos(angle), y = sin(angle), label_at = label_at)
  } else {
    data.frame(gene = character(), x = numeric(), y = numeric(),
               label_at = character())
  }

  rbind(centre_pos, ring_pos)
}

# Every segment of a plotly line trace shares one colour and width, so all the
# spokes of a view are drawn as a single trace with NA breaking them apart.
edge_path <- function(edges, pos) {
  if (nrow(edges) == 0) {
    return(list(x = NA_real_, y = NA_real_))
  }
  from <- pos[match(edges$target_gene, pos$gene), ]
  to   <- pos[match(edges$partner, pos$gene), ]
  list(
    x = as.vector(rbind(from$x, to$x, NA_real_)),
    y = as.vector(rbind(from$y, to$y, NA_real_))
  )
}

marker_size <- function(w, lo = 16, hi = 34) {
  if (length(w) == 0) return(numeric(0))
  span <- range(w)
  if (!is.finite(diff(span)) || diff(span) == 0) return(rep((lo + hi) / 2, length(w)))
  lo + (w - span[1]) / diff(span) * (hi - lo)
}

# star_positions() and edge_path() read target_gene and partner, so the DR
# table is renamed into that shape rather than the helpers being rewritten. A
# joint knockout puts every one of its genes at the centre, each spoked to
# every gene the knockout moved: a gene's movement belongs to the knockout as
# a whole, not to any one of its genes.
sctknk_edges <- sctknk_net %>%
  mutate(target_gene = target, partner = gene) %>%
  tidyr::separate_rows(target_gene, sep = ";")

views <- sctknk_edges %>%
  distinct(target, cell_type) %>%
  arrange(target, cell_type) %>%
  mutate(label = paste0(target, "  ·  ", cell_type))

fig <- plotly::plot_ly(width = NULL, height = 640)

# There is no legend: with no sign to tell edges apart every one of them is
# alike, so nothing is left for a legend to say. That also means the view
# traces start at 0 rather than after a pair of parked legend traces.
view_traces <- vector("list", nrow(views))
next_trace  <- 0L

for (i in seq_len(nrow(views))) {
  edges <- sctknk_edges %>%
    filter(target == views$target[i], cell_type == views$cell_type[i])

  pos <- star_positions(edges)
  visible <- i == 1

  spokes <- edge_path(edges, pos)

  hover <- paste0(
    edges$target_gene, " — ", edges$partner,
    "<br>log2FC ", signif(edges$log2FC, 4),
    "<br>p.adj ", signif(edges$p.adj, 3)
  )
  mid <- data.frame(
    x = (pos$x[match(edges$target_gene, pos$gene)] + pos$x[match(edges$partner, pos$gene)]) / 2,
    y = (pos$y[match(edges$target_gene, pos$gene)] + pos$y[match(edges$partner, pos$gene)]) / 2
  )

  # sctknk_sig holds a single row per cell type x target x gene, but a joint
  # knockout repeats each of them once per gene at the centre, so each node
  # is drawn from its first row.
  moved <- edges %>%
    distinct(partner, .keep_all = TRUE) %>%
    transmute(
      partner, log2FC,
      text = paste0(
        partner,
        "<br>log2FC ", signif(log2FC, 4),
        "<br>p.value ", signif(p.value, 3),
        "<br>p.adj ", signif(p.adj, 3)
      )
    ) %>%
    left_join(pos, by = c(partner = "gene"))

  centres <- pos[pos$gene %in% edges$target_gene, ]

  fig <- fig %>%
    plotly::add_trace(
      x = spokes$x, y = spokes$y, type = "scatter", mode = "lines",
      line = list(color = edge_colour, width = 2),
      hoverinfo = "skip", showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = mid$x, y = mid$y, type = "scatter", mode = "markers",
      marker = list(size = 14, color = "rgba(0,0,0,0)"),
      text = hover, hoverinfo = "text", showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = moved$x, y = moved$y, type = "scatter", mode = "markers+text",
      marker = list(
        size = marker_size(moved$log2FC),
        color = node_fill, opacity = 1,
        line = list(color = node_line, width = 1.5)
      ),
      text = moved$partner, textposition = moved$label_at,
      textfont = list(size = 11, color = "#0b0b0b"),
      hovertext = moved$text, hoverinfo = "text",
      showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = centres$x, y = centres$y, type = "scatter", mode = "markers+text",
      marker = list(
        size = 34, color = target_fill, opacity = 1,
        line = list(color = surface, width = 2)
      ),
      text = centres$gene, textposition = "bottom center",
      textfont = list(size = 13, color = "#0b0b0b"),
      hovertext = paste0(centres$gene, "<br>knocked out"), hoverinfo = "text",
      showlegend = FALSE, visible = visible
    )

  view_traces[[i]] <- next_trace + seq_len(4)
  next_trace <- next_trace + 4L
}

n_traces <- next_trace

# The title is restated on every button: a layout update replaces the title
# whole, so leaving out its placement would recentre it on the first switch.
title_style <- function(text) {
  list(text = text, x = 0, xanchor = "left",
       font = list(size = 15, color = "#0b0b0b"))
}

buttons <- lapply(seq_len(nrow(views)), function(i) {
  visible <- rep(FALSE, n_traces)
  visible[view_traces[[i]]] <- TRUE
  list(
    method = "update",
    args = list(
      list(visible = as.list(visible)),
      list(title = title_style(views$label[i]))
    ),
    label = views$label[i]
  )
})

fig %>%
  plotly::layout(
    title = title_style(views$label[1]),
    # No aspect lock: letting the ring stretch to the page width spreads the
    # gene labels apart instead of leaving the space either side empty.
    xaxis = list(visible = FALSE, range = c(-1.75, 1.75)),
    yaxis = list(visible = FALSE, range = c(-1.35, 1.35)),
    showlegend = FALSE,
    margin = list(l = 20, r = 20, t = 110, b = 40),
    paper_bgcolor = surface,
    plot_bgcolor = surface,
    hoverlabel = list(bgcolor = "#ffffff", bordercolor = "#d6d5d0"),
    updatemenus = list(list(
      type = "dropdown", direction = "down", active = 0,
      x = 1, xanchor = "right", y = 1.12, yanchor = "top",
      bgcolor = "#ffffff", bordercolor = "#d6d5d0",
      font = list(size = 12), buttons = buttons
    ))
  ) %>%
  plotly::config(displayModeBar = FALSE)
```

```{r sctknk-network-missing, eval=!(has_sctknk_net && has_plotly), results="asis"}
if (!has_sctknk) {
  cat("scTenifoldKnk was not run for this session, so there is no knockout",
      "network to draw. Pass `--sctknk` to generate one.")
} else if (!has_sctknk_sig) {
  cat("scTenifoldKnk ran, but no gene cleared FDR < 0.05 in any cell type x",
      "target combination, so there is nothing to draw.")
} else if (!has_sctknk_net) {
  cat("The only genes clearing FDR < 0.05 were the knocked-out genes",
      "themselves, which are excluded from the ring, so there is nothing to",
      "draw around them.")
} else {
  cat("plotly is not installed in this container, so the knockout networks",
      "could not be drawn. The table below still lists every gene.")
}
```

## Knockout differentially-regulated genes

Every gene clearing FDR < 0.05, across all cell type and target combinations and
with no cap on how many are listed per combination. `distance` is how far the
gene moved between the wild-type and knocked-out networks; `log2FC` is that
distance squared over the run's mean, logged; `p.value` tests it against the
chi-square null and `p.adj` is the Benjamini-Hochberg adjustment within the run.

`p.value` is a direct function of `log2FC`, so the two rank genes identically and
the FDR cut above is the same thing as a `log2FC` threshold. The knocked-out
genes themselves are left out: they always move furthest by construction, so
their rows say nothing about the knockout's effect.

Currently `r n_sctknk_sig` gene(s) clear FDR < 0.05, across
`r n_sctknk_targets` target(s) and `r n_sctknk_cell_types` cell type(s).

```{r sctknk-table, eval=has_sctknk_sig}
sctknk_sig %>%
  select(cell_type, target, gene, distance, log2FC, p.value, p.adj) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    # 0-based: column 5 is p.value
    options = list(pageLength = 15, order = list(list(5, "asc"))),
    caption = "scTenifoldKnk differentially-regulated genes at FDR < 0.05, per cell type x target."
  ) %>%
  formatSignif(columns = c("distance", "log2FC", "p.value", "p.adj"), digits = 4)
```

```{r sctknk-table-empty, eval=!has_sctknk_sig, results="asis"}
if (!has_sctknk) {
  cat("scTenifoldKnk was not run for this session, so no differentially-regulated",
      "gene table is shown. Pass `--sctknk` to generate one.")
} else {
  cat("scTenifoldKnk ran, but no gene cleared FDR < 0.05 in any cell type x",
      "target combination, so the table is empty.")
}
```


## Knockout gene set enrichment

What the knockout moved, read as processes rather than genes. Every gene is
ranked by `log2FC` -- the log2 of its squared manifold-alignment distance over
the run's mean squared distance, so the ranking is the one scTenifoldKnk's own
paper feeds to GSEA, centred on the average gene rather than on zero -- and each
gene set is scored on how concentrated it is at either end of that ranking. Use
the menu to switch between the `r n_gsea_views` target and cell type
combinations.

`NES` is the enrichment score normalised for set size, and it is signed here.
Read the sign carefully: the ranking is by how far a gene moved, not by which
way it moved, so a **positive** `NES` says a set is crowded into the genes the
knockout disturbed most, and a **negative** one says the set sat still while the
rest of the network moved. Neither is up- or down-regulation -- there is no
direction of expression anywhere in the scTenifoldKnk output to recover.

Each dot is one gene set: position on the x-axis is its `NES`, size is how many
genes of it sit in the leading edge, and colour is `-log(p.adj + 1)`, running
blue at the weak end of the cut through to red at the significant end. That
statistic is close to linear near `p.adj` of 0, so it separates the marginal
sets from the rest and treats everything below about 0.01 as equally
significant; the exact `p.adj` of every set is a column of the table. Sets are
drawn only if they clear FDR < `r gsea_fdr`, the `r gsea_top_terms` strongest
per combination by absolute `NES`. Hover any dot for its exact values; the
table below carries all of them, uncapped.

One thing to keep in mind reading any of it: the genes ranked here are
`gene4use`, the highly-variable, transcription-factor and drug-target set the
pipeline builds up front, not the whole transcriptome. Enrichment is relative to
that universe, which is already weighted toward regulators.

```{r gsea-figure, eval=has_gsea_fig && has_plotly}
ink       <- "#0b0b0b"
ink_muted <- "#52514e"
surface   <- "#fcfcfb"
gridline  <- "#e1e0d9"
baseline  <- "#c3c2b7"

# Blue at the weak end through to red at the significant end -- the convention
# an enrichment dot plot is read with. Both anchors and the neutral between
# them are the palette's own blue/red pair; the midpoint is that pair's grey
# taken a couple of steps down from the surface, because the documented
# near-white midpoint would put a dot where a reader sees nothing at all.
dot_scale <- list(
  list(0.0, "#2a78d6"),
  list(0.5, "#c3c2b7"),
  list(1.0, "#e34948")
)

views <- gsea_fig %>%
  distinct(target, cell_type) %>%
  arrange(target, cell_type)

# Size, colour and the x-axis are all scaled once over every view rather than
# per view. The menu swaps traces in place, so a scale that moved with it would
# make two dots of the same size, shade or position mean different things one
# click apart -- the axis would rescale under a reader who is holding the
# previous view in their head.
size_max_px <- 26
size_ref    <- 2 * max(gsea_fig$leading_edge_n, na.rm = TRUE) / size_max_px^2

# The colour range is the whole span the FDR cut can produce -- p.adj just
# under gsea_fdr at one end, p.adj approaching 0 at the other -- rather than
# the range this run happens to hold. Derived rather than measured, so a shade
# means the same thing in a report with three significant sets as in one with
# three hundred, and nothing can fall off either end of the bar.
colour_min   <- -log(1 + gsea_fdr)
colour_max   <- 0
colour_ticks <- pretty(c(colour_min, colour_max), n = 4)
colour_ticks <- colour_ticks[colour_ticks >= colour_min & colour_ticks <= colour_max]
colour_labels <- format(colour_ticks, trim = TRUE)

# Always includes zero: the sign of NES is the whole point of the axis, so the
# baseline it is signed against has to be on screen even when every set in a
# view happens to land on one side of it.
x_span <- range(c(0, gsea_fig$NES), na.rm = TRUE)
x_pad  <- max(diff(x_span) * 0.08, 0.1)
x_lim  <- c(x_span[1] - x_pad, x_span[2] + x_pad)

# Dots sit at numeric y positions with the labels supplied per view through the
# menu, rather than on a categorical axis. A categorical axis is shared by every
# trace on the figure, so switching views would leave the sets of all the other
# combinations on it as empty rows.
view_rows <- vector("list", nrow(views))
fig <- plotly::plot_ly()

for (i in seq_len(nrow(views))) {
  d <- gsea_fig %>%
    filter(target == views$target[i], cell_type == views$cell_type[i]) %>%
    arrange(NES)

  view_rows[[i]] <- d

  hover <- paste0(
    d$pathway,
    "<br>NES ", signif(d$NES, 3),
    "<br>p.adj ", signif(d$padj, 3),
    "<br>leading edge ", d$leading_edge_n, " of ", d$size, " gene(s)"
  )

  fig <- fig %>%
    plotly::add_trace(
      x = d$NES, y = seq_len(nrow(d)),
      type = "scatter", mode = "markers",
      marker = list(
        size     = d$leading_edge_n,
        sizemode = "area",
        sizeref  = size_ref,
        # A set can carry a single leading-edge gene, which at this sizeref is
        # a dot too small to hover, let alone see.
        sizemin  = 8,
        color      = d$neg_log_padj,
        colorscale = dot_scale,
        cmin = colour_min, cmax = colour_max,
        # The ring is surface-coloured rather than a border: it keeps two dots
        # that land on the same NES legible without adding ink that isn't data.
        line = list(color = surface, width = 2),
        # Every trace carries the scale; plotly draws the bar only for the one
        # that is visible, so it survives a view switch unchanged.
        showscale = TRUE,
        colorbar = list(
          title = list(text = "-log(p.adj + 1)", side = "top",
                       font = list(size = 11, color = ink_muted)),
          tickvals = colour_ticks, ticktext = colour_labels,
          tickfont = list(size = 11, color = ink_muted),
          # Pixels, not a fraction of the plot: the views are different
          # heights, and a fractional bar would be a squashed stub with its
          # ticks on top of each other on a combination that found three sets.
          thickness = 10, lenmode = "pixels", len = 160, outlinewidth = 0,
          x = 1.02, xanchor = "left", y = 1, yanchor = "top"
        )
      ),
      hovertext = hover, hoverinfo = "text",
      showlegend = FALSE, visible = i == 1
    )
}

# The axis is rebuilt on every switch, so its whole state has to be restated in
# each button -- the same reason the title is restated in the knockout figure.
view_layout <- function(i) {
  d <- view_rows[[i]]
  n <- nrow(d)

  list(
    title = list(
      text = paste0(d$target[1], " knocked out in ", d$cell_type[1]),
      x = 0, xanchor = "left",
      font = list(size = 15, color = ink)
    ),
    # Height travels with the view rather than being fixed once for the
    # figure. One slot per set at a constant 34px means a combination with two
    # sets gets a short figure with normal-sized dots, instead of two dots
    # marooned in the height a twenty-set view needs.
    height = 34 * n + 150,
    yaxis = list(
      tickvals = seq_len(n),
      ticktext = pretty_set(d$pathway),
      # Half a slot of air top and bottom, which is 17px against the 34px
      # slot below -- more than the radius of the largest dot, so a big dot on
      # the first or last row is not clipped by the plot edge.
      range = c(0.45, n + 0.55),
      tickfont = list(size = 11, color = ink_muted),
      # One hairline per row, so a dot 300px from its label stays on the right
      # one. Solid, a single step off the surface: it has to be followable
      # without competing with the dots.
      showgrid = TRUE, gridcolor = gridline, gridwidth = 1,
      zeroline = FALSE, showline = FALSE, title = ""
    ),
    xaxis = list(
      range = x_lim,
      title = list(text = "NES", font = list(size = 12, color = ink_muted)),
      tickfont = list(size = 11, color = ink_muted),
      gridcolor = gridline,
      # The one rule on the plot drawn heavier than a gridline: it is what the
      # sign of every dot is read against.
      zeroline = TRUE, zerolinecolor = baseline, zerolinewidth = 2,
      showline = FALSE
    )
  )
}

buttons <- lapply(seq_len(nrow(views)), function(i) {
  visible <- rep(FALSE, nrow(views))
  visible[i] <- TRUE

  list(
    method = "update",
    args = list(list(visible = visible), view_layout(i)),
    label = paste0(views$target[i], " - ", views$cell_type[i])
  )
})

first <- view_layout(1)

fig %>%
  plotly::layout(
    title  = first$title,
    xaxis  = first$xaxis,
    yaxis  = first$yaxis,
    height = first$height,
    margin = list(l = 330, r = 180, t = 90, b = 55),
    paper_bgcolor = surface,
    plot_bgcolor  = surface,
    updatemenus = list(list(
      active = 0, buttons = buttons,
      # Anchored by its bottom edge, so the menu grows up into the top
      # margin instead of down over the first row of dots on a combination
      # that only found two or three sets.
      x = 1, xanchor = "right", y = 1.16, yanchor = "bottom",
      direction = "down", showactive = TRUE,
      bgcolor = surface, bordercolor = "#d9d8d3"
    ))
  ) %>%
  plotly::config(displayModeBar = FALSE)
```

```{r gsea-figure-missing, eval=!(has_gsea_fig && has_plotly), results="asis"}
if (!gsea_requested) {
  cat("No gene set file was supplied, so no enrichment was run. Pass",
      "`--gsea_gmt` with a GMT for this run's species alongside `--sctknk`.")
} else if (!has_gsea) {
  cat("The enrichment ran and returned nothing. The commonest cause is a gene",
      "set file whose symbols do not match the data's own -- a human GMT",
      "against mouse symbols matches nothing. The run log names the symbols on",
      "each side and counts the overlap.")
} else if (!has_gsea_sig) {
  cat("The enrichment ran, but no gene set cleared FDR <", gsea_fdr, "in any",
      "cell type x target combination, so there is nothing to draw.")
} else if (!has_gsea_fig) {
  cat("Every significant gene set came back without a usable NES or p.adj, so",
      "there is nothing to plot. The table below still lists them.")
} else {
  cat("plotly is not installed in this container, so the enrichment figure",
      "could not be drawn. The table below still lists every gene set.")
}
```

Every gene set clearing FDR < `r gsea_fdr`, uncapped -- the figure above shows
only the strongest `r gsea_top_terms` of each combination. `leading_edge_n` is
how many of the set's genes sit in the leading edge, the stretch of the ranking
that carries the set's score, and it is what the figure sizes each dot by. The
leading edge's gene names are not repeated here; they are in
`sctknk/gsea_all_targets.txt` alongside `ES` and the unadjusted `pval`. Set
names longer than 50 characters are shown cut, with the whole name on hover --
the column filter still searches the part that is not shown.

```{r gsea-table, eval=has_gsea_sig}
gsea_sig %>%
  select(cell_type, target, pathway, NES, size, leading_edge_n, padj) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    # Only the pathway column renders as HTML, and only because the truncation
    # below builds a span for it.
    escape = -3,
    options = list(
      pageLength = 15,
      # 0-based: column 6 is padj. pval used to carry this sort; it is a
      # monotone function of padj within a combination, so the order the reader
      # sees is unchanged by dropping the column.
      order = list(list(6, "asc")),
      # Set names arrive in MSigDB's shape -- GOBP_REGULATION_OF_... -- and a
      # collection's longest run past 120 characters, which widens the column
      # far enough to push p.adj off a laptop screen. Cut at the same 50 the
      # figure's axis labels use. Truncating in the renderer rather than in the
      # data keeps the whole name in the cell, so the column filter still
      # matches on the part that is not shown and the tooltip can show it.
      columnDefs = list(list(
        targets = 2,
        render = DT::JS(
          "function(data, type, row) {",
          "  if (type !== 'display' || data === null) return data;",
          "  var text = String(data);",
          "  if (text.length <= 50) return text;",
          "  return '<span title=\"' + text + '\">' +",
          "         text.substr(0, 49) + '\u2026</span>';",
          "}"
        )
      ))
    ),
    caption = paste0("Gene sets enriched at either end of each knockout's ",
                     "ranking, FDR < ", gsea_fdr, ".")
  ) %>%
  formatSignif(columns = c("NES", "padj"), digits = 4)
```

```{r gsea-table-empty, eval=!has_gsea_sig, results="asis"}
if (!gsea_requested) {
  cat("No gene set file was supplied, so no enrichment was run and there is no",
      "table to show. Pass `--gsea_gmt` alongside `--sctknk`.")
} else if (!has_gsea) {
  cat("The enrichment ran and returned nothing; see the note above the figure.")
} else {
  cat("The enrichment ran, but no gene set cleared FDR <", gsea_fdr, "in any",
      "cell type x target combination, so the table is empty.")
}
```


## Cell identity overview

The identities scored above, as they sit in the embedding. This is the cell set
after downsampling, so the group sizes are the ones the networks were actually
inferred from rather than those of the full object.

```{r umap, eval=has_umap, out.width="100%", fig.align="center"}
knitr::include_graphics(umap_path)
```

```{r umap-missing, eval=!has_umap, results="asis"}
cat("No UMAP figure was supplied with this report.")
```
