---
title: "NetPerturb Report"
subtitle: "Perturbation scores across targets and cell types"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    number-sections: true
    page-layout: full
    theme: flatly
params:
  perbscore_file: "perbscore_all_targets.txt"
  umap_file: "umap.png"
  network: ""
  score_quantile: 0.75
  sctknk_file: ""
  sctknk_top_genes: 25
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(scales)

# plotly draws the knockout networks. A container built before that section
# existed still has to render the rest of the report, so its absence is a
# missing figure rather than an error.
has_plotly <- requireNamespace("plotly", quietly = TRUE)

perb_score <- read.delim(params$perbscore_file, header = TRUE, sep = "\t")

n_cell_types <- n_distinct(perb_score$cell_type)
n_targets    <- n_distinct(perb_score$target)
n_bindings   <- n_distinct(perb_score$binding)

# Pool every score from every target on a log10 scale; non-positive scores
# cannot be logged and are carried as NA.
score_data <- perb_score %>%
  mutate(log_score = ifelse(perb_score > 0, log10(perb_score), NA_real_))

has_scores <- any(is.finite(score_data$log_score))

# Pooled scores are bimodal: a low mode of near-zero scores from genes the
# network barely leans on, and a high mode that carries the signal. Imputing
# the low mode up to a bound only stacked it against that bound and left it in
# the picture, so it is cut instead. Everything below the score_quantile
# quantile of the pooled log10 scores is dropped and the high mode is plotted
# on its own; the default 0.75 cuts at q3.
score_quantile <- suppressWarnings(as.numeric(params$score_quantile))
if (!isTRUE(is.finite(score_quantile)) || score_quantile < 0 || score_quantile >= 1) {
  score_quantile <- 0.75
}

score_cut <- if (has_scores) {
  unname(quantile(score_data$log_score, score_quantile, na.rm = TRUE))
} else {
  NA_real_
}

score_data <- score_data %>%
  mutate(log_score_kept = ifelse(
    is.finite(log_score) & log_score >= score_cut, log_score, NA_real_
  ))

n_scored  <- sum(is.finite(score_data$log_score))
n_kept    <- sum(is.finite(score_data$log_score_kept))
n_below   <- n_scored - n_kept
n_dropped <- sum(!is.finite(score_data$log_score))

# The strongest scores in the run. "Overall" is the whole pooled set, the same
# population the retention cut is taken from, so this threshold does not move
# with score_quantile. Change top_quantile alone to mark a different fraction.
top_quantile <- 0.975
top_label    <- paste0("Top ", signif((1 - top_quantile) * 100, 3), "%")

top_cut <- if (has_scores) {
  unname(quantile(score_data$log_score, top_quantile, na.rm = TRUE))
} else {
  NA_real_
}

score_data <- score_data %>%
  mutate(is_top = is.finite(log_score) & log_score >= top_cut)

n_top <- sum(score_data$is_top)

# The network inference method the run was launched with (--network). The
# report also has to render when it is not supplied, e.g. when the qmd is
# knitted by hand against a table someone already has.
network_key   <- if (is.null(params$network)) "" else trimws(as.character(params$network))
network_names <- c(
  genie3  = "GENIE3",
  scrank  = "scRank",
  hdwgcna = "hdWGCNA"
)
network_label <- unname(network_names[network_key])
if (!nzchar(network_key)) {
  network_label <- "an unreported method"
} else if (is.na(network_label)) {
  network_label <- network_key
}
network_value <- if (nzchar(network_key)) network_key else "not supplied"

# The UMAP is a QC figure produced upstream by DOWNSAMPLE. A run started before
# that step existed, or one whose plot could not be drawn, still has to render.
umap_path <- params$umap_file
has_umap  <- !is.null(umap_path) && nzchar(umap_path) && file.exists(umap_path)
if (has_umap) umap_path <- normalizePath(umap_path)

# scTenifoldKnk's differentially-regulated (DR) gene table, from a --sctknk
# run. A header-only sentinel file (no rows) stands in when it was not
# requested for this run at all, same shape as a real run that just found
# nothing, so both render the same "nothing to show" message.
# The named-but-empty frame matters: everything below filters on p.adj, so a
# bare data.frame() with no columns would fail with "object 'p.adj' not found"
# rather than fall through to the "not run" message. Same reasoning, and the
# same 8 columns, as empty_dr in bin/sctenifoldknk.R. The size check is for
# MERGE_SCTENIFOLDKNK's stub, which touches a zero-byte file that read.delim
# refuses with "no lines available in input".
sctknk_empty <- data.frame(
  cell_type = character(), target = character(), gene = character(),
  distance = numeric(), Z = numeric(), FC = numeric(),
  p.value = numeric(), p.adj = numeric(), stringsAsFactors = FALSE
)

sctknk_path <- params$sctknk_file
sctknk_dr <- if (!is.null(sctknk_path) && nzchar(sctknk_path) &&
                  file.exists(sctknk_path) && file.size(sctknk_path) > 0) {
  read.delim(sctknk_path, header = TRUE, sep = "\t")
} else {
  sctknk_empty
}

has_sctknk <- nrow(sctknk_dr) > 0

# How many genes the network figure draws around each target. The table below
# it is never capped.
sctknk_top_genes <- suppressWarnings(as.integer(params$sctknk_top_genes))
if (!isTRUE(is.finite(sctknk_top_genes)) || !isTRUE(sctknk_top_genes >= 1)) {
  sctknk_top_genes <- 25L
}

# dRegulation() reports FC as a gene's squared manifold-alignment distance over
# the mean squared distance of the run, so it is a ratio to the average rather
# than a differential-expression fold change, and it carries no direction: a
# gene that moved a long way scores high whichever way it moved. log2 of it is
# what the figure sizes nodes by and what the table reports.
if (has_sctknk) {
  sctknk_dr <- sctknk_dr %>%
    mutate(log2FC = ifelse(is.finite(FC) & FC > 0, log2(FC), NA_real_))
} else {
  # No rows, but the 8 columns are there either way, so only log2FC has to be
  # added to keep the select() in the table below working.
  sctknk_dr$log2FC <- numeric()
}

# p.value is pchisq(FC, df = 1) and p.adj is BH over that, so the FDR cut is a
# monotone function of FC: filtering here is the same as taking everything
# above a log2FC threshold. Both sections below start from this frame.
sctknk_sig <- sctknk_dr %>%
  filter(!is.na(p.adj), p.adj < 0.05) %>%
  arrange(target, cell_type, p.value)

has_sctknk_sig <- nrow(sctknk_sig) > 0
n_sctknk_sig   <- nrow(sctknk_sig)

# Counted over the significant rows, not over everything knocked out: a target
# that came back with nothing would otherwise be counted in a sentence that
# says how far the hits reach.
n_sctknk_targets    <- n_distinct(sctknk_sig$target)
n_sctknk_cell_types <- n_distinct(sctknk_sig$cell_type)

# Ring nodes for the figure. The knocked-out gene is always in its own DR table
# and always moves furthest, so it would be drawn as a spoke back to itself;
# it is dropped here and kept in the table, where it reads as confirmation the
# knockout took. The ring is then capped, since a target can clear FDR on
# hundreds of genes and every view is embedded in the HTML up front.
sctknk_net <- if (has_sctknk_sig) {
  sctknk_sig %>%
    filter(gene != target, is.finite(log2FC)) %>%
    group_by(target, cell_type) %>%
    slice_max(log2FC, n = sctknk_top_genes, with_ties = FALSE) %>%
    ungroup()
} else {
  sctknk_sig
}

has_sctknk_net <- nrow(sctknk_net) > 0

# Counted off sctknk_net rather than sctknk_sig, because that is the frame the
# dropdown is built from: a combination whose only significant gene was the
# knocked-out gene itself has a row in the table but no menu entry.
n_sctknk_views <- if (has_sctknk_net) {
  nrow(distinct(sctknk_net, target, cell_type))
} else {
  0
}
```

## Overview

This report summarizes `r nrow(perb_score)` perturbation scores from
`r params$perbscore_file`, covering `r n_targets` target(s) across
`r n_cell_types` cell type(s), scored under `r n_bindings` binding
mode(s) (`r paste(sort(unique(perb_score$binding)), collapse = ", ")`).

The gene regulatory networks these scores were computed from were inferred with
**`r network_label`** (`--network`: `r network_value`). Every score below is
relative to that network type, so scores from a run using a different inference
method are not directly comparable.

## Perturbation score table

Use the column filters below to query by cell type, target, or binding.

```{r table}
perb_score %>%
  arrange(target, cell_type) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    options = list(pageLength = 15, order = list(list(1, "asc"))),
    caption = "All target x cell type perturbation scores."
  ) %>%
  formatSignif(columns = "perb_score", digits = 4)
```

## Perturbation score distribution

All `r n_scored` log10 scores from every target and cell type pooled into one
distribution. Pooled scores are typically bimodal: a low mode of near-zero
values from genes the network barely leans on, and a high mode carrying the
signal. Only the high mode is kept, cut at the `r score_quantile` quantile
(red line, `r if (has_scores) signif(score_cut, 4) else NA`). That retains
`r n_kept` score(s) and drops `r n_below` below the cut; a further `r n_dropped`
non-positive score(s) cannot be logged and are not shown either.

The left panel is every score, so the cut can be checked against where the two
modes actually separate — move it with `--score_quantile` if the red line does
not fall in the valley between them. The right panel is what the heatmap below
works with, on its own axis.

The green dashed line in both panels is the `r top_label` threshold
(`r if (has_scores) signif(top_cut, 4) else NA`), the `r n_top` strongest
score(s) in the run. Those are the scores the heatmap asterisks mark.

```{r distribution, fig.width=10, fig.height=4, eval=has_scores}
dist_data <- score_data %>%
  select(log_score, log_score_kept) %>%
  pivot_longer(everything(), names_to = "stage", values_to = "value") %>%
  filter(is.finite(value)) %>%
  mutate(stage = factor(
    stage,
    levels = c("log_score", "log_score_kept"),
    labels = c("All scores", "Retained high mode")
  ))

# The retention cut belongs only to the panel it acts on; the top threshold is
# drawn in both, so the strongest scores can be placed against the whole set
# and against the mode the report keeps.
stages <- levels(dist_data$stage)

ref_lines <- rbind(
  data.frame(
    stage = factor("All scores", levels = stages),
    xintercept = score_cut,
    line = "Quantile cut"
  ),
  data.frame(
    stage = factor(stages, levels = stages),
    xintercept = top_cut,
    line = top_label
  )
)

# Red and green are a poor pair for the commonest colour blindness, so the two
# lines differ in dash as well as hue and the legend names both.
ref_colours   <- setNames(c("#e34948", "#008300"), c("Quantile cut", top_label))
ref_linetypes <- setNames(c("solid", "dashed"), c("Quantile cut", top_label))

# A density curve is what makes the two modes readable, but it needs a few
# points per panel to mean anything.
show_density <- min(table(dist_data$stage)) >= 3

ggplot(dist_data, aes(x = value)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 20, fill = "steelblue", color = "white"
  ) +
  { if (show_density) geom_density(color = "grey25", linewidth = 0.6) } +
  geom_vline(
    data = ref_lines,
    aes(xintercept = xintercept, colour = line, linetype = line),
    linewidth = 0.7
  ) +
  scale_colour_manual(name = NULL, values = ref_colours) +
  scale_linetype_manual(name = NULL, values = ref_linetypes) +
  facet_wrap(~stage, scales = "free") +
  labs(
    title = "Distribution of perturbation scores across all targets",
    x = "log10(perturbation score)",
    y = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r distribution-empty, eval=!has_scores, results="asis"}
cat("No positive perturbation score is available, so no distribution is shown.")
```

## Perturbation score heatmap

Scores are log10-transformed and cut as above, so the colour scale spans the
high mode only. Tiles below the cut, and those whose score is non-positive, are
shown as missing (gray). An asterisk marks a score in the `r top_label` of every
score in the run — `r n_top` tile(s), at or above
`r if (has_scores) signif(top_cut, 4) else NA` on the log10 scale.

```{r heatmap, fig.width=10, fig.height=max(4, 0.4 * n_cell_types)}
ggplot(score_data, aes(x = target, y = cell_type, fill = log_score_kept)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(
    data = filter(score_data, is_top),
    aes(label = "*"), colour = "#0b0b0b", size = 6, vjust = 0.75
  ) +
  scale_fill_viridis_c(name = "log10(perb_score)", na.value = "gray80") +
  labs(x = "Target", y = "Cell type") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  { if (n_bindings > 1) facet_wrap(~binding) }
```

## Knockout networks

Each knocked-out gene and the genes its knockout moved. scTenifoldKnk zeroes the
target's outgoing edges, realigns the knocked-out network against the wild-type
one, and scores every gene on how far it travelled between the two; the genes
drawn here are the ones clearing FDR < 0.05. Use the menu to switch between the
`r n_sctknk_views` target and cell type combinations.

The knocked-out gene sits at the centre and the genes it moved sit around it. A
node's size follows its `log2FC`, scaled within the combination on display, so
sizes rank genes against each other rather than across cell types. The ring is
capped at the `r sctknk_top_genes` strongest genes per combination; the table in
the next section is not capped. Hover any node for its exact values.

`log2FC` here is the log2 ratio of a gene's squared alignment distance to the
mean squared distance of its run, so it says how far a gene moved and not which
way it went. It carries no direction, which is why no edge on this figure has a
sign to read off: every gene shown simply moved further than its run's average.

```{r sctknk-network, eval=has_sctknk_net && has_plotly}
# log2FC has no sign and is already carried by node size, so nothing on this
# figure needs to carry meaning through colour and the palette stays neutral.
edge_colour <- "#8f8e8a"
node_fill   <- "#e8e7e3"
node_line   <- "#52514e"
target_fill <- "#3f3e3b"
surface     <- "#fcfcfb"

# The subnetwork of one knockout is a star, so it needs no force layout: the
# knocked-out gene goes in the middle and the genes it moved onto a ring.
star_positions <- function(edges) {
  centre <- sort(unique(edges$target_gene))
  ring   <- setdiff(unique(edges$partner), centre)

  centre_pos <- if (length(centre) == 1) {
    data.frame(gene = centre, x = 0, y = 0, label_at = "bottom center")
  } else {
    angle <- (seq_along(centre) - 1) * 2 * pi / length(centre) + pi / 2
    data.frame(gene = centre, x = 0.22 * cos(angle), y = 0.22 * sin(angle),
               label_at = "bottom center")
  }

  ring_pos <- if (length(ring) > 0) {
    angle <- (seq_along(ring) - 1) * 2 * pi / length(ring) + pi / 2
    # A label pushed straight out from the centre clears both its own node and
    # the edge running into it, which "top center" on every node does not.
    label_at <- ifelse(
      cos(angle) > 0.5, "middle right",
      ifelse(cos(angle) < -0.5, "middle left",
             ifelse(sin(angle) > 0, "top center", "bottom center"))
    )
    data.frame(gene = ring, x = cos(angle), y = sin(angle), label_at = label_at)
  } else {
    data.frame(gene = character(), x = numeric(), y = numeric(),
               label_at = character())
  }

  rbind(centre_pos, ring_pos)
}

# Every segment of a plotly line trace shares one colour and width, so all the
# spokes of a view are drawn as a single trace with NA breaking them apart.
edge_path <- function(edges, pos) {
  if (nrow(edges) == 0) {
    return(list(x = NA_real_, y = NA_real_))
  }
  from <- pos[match(edges$target_gene, pos$gene), ]
  to   <- pos[match(edges$partner, pos$gene), ]
  list(
    x = as.vector(rbind(from$x, to$x, NA_real_)),
    y = as.vector(rbind(from$y, to$y, NA_real_))
  )
}

marker_size <- function(w, lo = 16, hi = 34) {
  if (length(w) == 0) return(numeric(0))
  span <- range(w)
  if (!is.finite(diff(span)) || diff(span) == 0) return(rep((lo + hi) / 2, length(w)))
  lo + (w - span[1]) / diff(span) * (hi - lo)
}

# star_positions() and edge_path() read target_gene and partner, so the DR
# table is renamed into that shape rather than the helpers being rewritten.
sctknk_edges <- sctknk_net %>%
  mutate(target_gene = target, partner = gene)

views <- sctknk_edges %>%
  distinct(target, cell_type) %>%
  arrange(target, cell_type) %>%
  mutate(label = paste0(target, "  ·  ", cell_type))

fig <- plotly::plot_ly(width = NULL, height = 640)

# There is no legend: with no sign to tell edges apart every one of them is
# alike, so nothing is left for a legend to say. That also means the view
# traces start at 0 rather than after a pair of parked legend traces.
view_traces <- vector("list", nrow(views))
next_trace  <- 0L

for (i in seq_len(nrow(views))) {
  edges <- sctknk_edges %>%
    filter(target == views$target[i], cell_type == views$cell_type[i])

  pos <- star_positions(edges)
  visible <- i == 1

  spokes <- edge_path(edges, pos)

  hover <- paste0(
    edges$target_gene, " — ", edges$partner,
    "<br>log2FC ", signif(edges$log2FC, 4),
    "<br>p.adj ", signif(edges$p.adj, 3)
  )
  mid <- data.frame(
    x = (pos$x[match(edges$target_gene, pos$gene)] + pos$x[match(edges$partner, pos$gene)]) / 2,
    y = (pos$y[match(edges$target_gene, pos$gene)] + pos$y[match(edges$partner, pos$gene)]) / 2
  )

  # One row per gene already: sctknk_sig holds a single row per cell type x
  # target x gene, so no grouping is needed to draw each node once.
  moved <- edges %>%
    transmute(
      partner, log2FC,
      text = paste0(
        partner,
        "<br>log2FC ", signif(log2FC, 4),
        "<br>p.value ", signif(p.value, 3),
        "<br>p.adj ", signif(p.adj, 3)
      )
    ) %>%
    left_join(pos, by = c(partner = "gene"))

  centres <- pos[pos$gene %in% edges$target_gene, ]

  fig <- fig %>%
    plotly::add_trace(
      x = spokes$x, y = spokes$y, type = "scatter", mode = "lines",
      line = list(color = edge_colour, width = 2),
      hoverinfo = "skip", showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = mid$x, y = mid$y, type = "scatter", mode = "markers",
      marker = list(size = 14, color = "rgba(0,0,0,0)"),
      text = hover, hoverinfo = "text", showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = moved$x, y = moved$y, type = "scatter", mode = "markers+text",
      marker = list(
        size = marker_size(moved$log2FC),
        color = node_fill, opacity = 1,
        line = list(color = node_line, width = 1.5)
      ),
      text = moved$partner, textposition = moved$label_at,
      textfont = list(size = 11, color = "#0b0b0b"),
      hovertext = moved$text, hoverinfo = "text",
      showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = centres$x, y = centres$y, type = "scatter", mode = "markers+text",
      marker = list(
        size = 34, color = target_fill, opacity = 1,
        line = list(color = surface, width = 2)
      ),
      text = centres$gene, textposition = "bottom center",
      textfont = list(size = 13, color = "#0b0b0b"),
      hovertext = paste0(centres$gene, "<br>knocked out"), hoverinfo = "text",
      showlegend = FALSE, visible = visible
    )

  view_traces[[i]] <- next_trace + seq_len(4)
  next_trace <- next_trace + 4L
}

n_traces <- next_trace

# The title is restated on every button: a layout update replaces the title
# whole, so leaving out its placement would recentre it on the first switch.
title_style <- function(text) {
  list(text = text, x = 0, xanchor = "left",
       font = list(size = 15, color = "#0b0b0b"))
}

buttons <- lapply(seq_len(nrow(views)), function(i) {
  visible <- rep(FALSE, n_traces)
  visible[view_traces[[i]]] <- TRUE
  list(
    method = "update",
    args = list(
      list(visible = as.list(visible)),
      list(title = title_style(views$label[i]))
    ),
    label = views$label[i]
  )
})

fig %>%
  plotly::layout(
    title = title_style(views$label[1]),
    # No aspect lock: letting the ring stretch to the page width spreads the
    # gene labels apart instead of leaving the space either side empty.
    xaxis = list(visible = FALSE, range = c(-1.75, 1.75)),
    yaxis = list(visible = FALSE, range = c(-1.35, 1.35)),
    showlegend = FALSE,
    margin = list(l = 20, r = 20, t = 110, b = 40),
    paper_bgcolor = surface,
    plot_bgcolor = surface,
    hoverlabel = list(bgcolor = "#ffffff", bordercolor = "#d6d5d0"),
    updatemenus = list(list(
      type = "dropdown", direction = "down", active = 0,
      x = 1, xanchor = "right", y = 1.12, yanchor = "top",
      bgcolor = "#ffffff", bordercolor = "#d6d5d0",
      font = list(size = 12), buttons = buttons
    ))
  ) %>%
  plotly::config(displayModeBar = FALSE)
```

```{r sctknk-network-missing, eval=!(has_sctknk_net && has_plotly), results="asis"}
if (!has_sctknk) {
  cat("scTenifoldKnk was not run for this session, so there is no knockout",
      "network to draw. Pass `--sctknk` to generate one.")
} else if (!has_sctknk_sig) {
  cat("scTenifoldKnk ran, but no gene cleared FDR < 0.05 in any cell type x",
      "target combination, so there is nothing to draw.")
} else if (!has_sctknk_net) {
  cat("The only genes clearing FDR < 0.05 were the knocked-out genes",
      "themselves, which are excluded from the ring, so there is nothing to",
      "draw around them.")
} else {
  cat("plotly is not installed in this container, so the knockout networks",
      "could not be drawn. The table below still lists every gene.")
}
```

## Knockout differentially-regulated genes

Every gene clearing FDR < 0.05, across all cell type and target combinations and
with no cap on how many are listed per combination. `distance` is how far the
gene moved between the wild-type and knocked-out networks; `log2FC` is that
distance squared over the run's mean, logged; `p.value` tests it against the
chi-square null and `p.adj` is the Benjamini-Hochberg adjustment within the run.

`p.value` is a direct function of `log2FC`, so the two rank genes identically and
the FDR cut above is the same thing as a `log2FC` threshold. Each knocked-out
gene appears in its own rows, normally at the top, which is the simplest check
that the knockout took.

Currently `r n_sctknk_sig` gene(s) clear FDR < 0.05, across
`r n_sctknk_targets` target(s) and `r n_sctknk_cell_types` cell type(s).

```{r sctknk-table, eval=has_sctknk_sig}
sctknk_sig %>%
  select(cell_type, target, gene, distance, log2FC, p.value, p.adj) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    # 0-based: column 5 is p.value
    options = list(pageLength = 15, order = list(list(5, "asc"))),
    caption = "scTenifoldKnk differentially-regulated genes at FDR < 0.05, per cell type x target."
  ) %>%
  formatSignif(columns = c("distance", "log2FC", "p.value", "p.adj"), digits = 4)
```

```{r sctknk-table-empty, eval=!has_sctknk_sig, results="asis"}
if (!has_sctknk) {
  cat("scTenifoldKnk was not run for this session, so no differentially-regulated",
      "gene table is shown. Pass `--sctknk` to generate one.")
} else {
  cat("scTenifoldKnk ran, but no gene cleared FDR < 0.05 in any cell type x",
      "target combination, so the table is empty.")
}
```


## Cell identity overview

The identities scored above, as they sit in the embedding. This is the cell set
after downsampling, so the group sizes are the ones the networks were actually
inferred from rather than those of the full object.

```{r umap, eval=has_umap, out.width="100%", fig.align="center"}
knitr::include_graphics(umap_path)
```

```{r umap-missing, eval=!has_umap, results="asis"}
cat("No UMAP figure was supplied with this report.")
```
