---
title: "NetPerturb Report"
subtitle: "Perturbation scores across targets and cell types"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    number-sections: true
    page-layout: full
    theme: flatly
params:
  perbscore_file: "perbscore_all_targets.txt"
  umap_file: "umap.png"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(scales)

perb_score <- read.delim(params$perbscore_file, header = TRUE, sep = "\t")

n_cell_types <- n_distinct(perb_score$cell_type)
n_targets    <- n_distinct(perb_score$target)
n_bindings   <- n_distinct(perb_score$binding)

# Pool every score from every target on a log10 scale; non-positive scores
# cannot be logged and are carried as NA.
score_data <- perb_score %>%
  mutate(log_score = ifelse(perb_score > 0, log10(perb_score), NA_real_))

has_scores <- any(is.finite(score_data$log_score))

# Quantile cut: values below q1 - 1.5 * IQR are low-expression outliers that
# squash the colour scale, so they are imputed up to that bound.
if (has_scores) {
  q1  <- quantile(score_data$log_score, 0.25, na.rm = TRUE)
  q3  <- quantile(score_data$log_score, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- unname(q1 - 1.5 * iqr)
} else {
  lower_bound <- NA_real_
}

score_data <- score_data %>%
  mutate(log_score_imputed = pmax(log_score, lower_bound))

n_imputed <- sum(score_data$log_score < lower_bound, na.rm = TRUE)
n_dropped <- sum(!is.finite(score_data$log_score))

# The UMAP is a QC figure produced upstream by DOWNSAMPLE. A run started before
# that step existed, or one whose plot could not be drawn, still has to render.
umap_path <- params$umap_file
has_umap  <- !is.null(umap_path) && nzchar(umap_path) && file.exists(umap_path)
if (has_umap) umap_path <- normalizePath(umap_path)
```

## Overview

This report summarizes `r nrow(perb_score)` perturbation scores from
`r params$perbscore_file`, covering `r n_targets` target(s) across
`r n_cell_types` cell type(s), scored under `r n_bindings` binding
mode(s) (`r paste(sort(unique(perb_score$binding)), collapse = ", ")`).

## Perturbation score table

Use the column filters below to query by cell type, target, or binding.

```{r table}
perb_score %>%
  arrange(target, cell_type) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    options = list(pageLength = 15, order = list(list(1, "asc"))),
    caption = "All target x cell type perturbation scores."
  ) %>%
  formatSignif(columns = "perb_score", digits = 4)
```

## Perturbation score distribution

All `r sum(is.finite(score_data$log_score))` log10 scores from every target and
cell type pooled into one distribution. To avoid low-expression biases, values
below the quantile cut `q1 - 1.5 * IQR` (red line,
`r if (has_scores) signif(lower_bound, 4) else NA`) are imputed up to that bound:
`r n_imputed` value(s) were imputed and `r n_dropped` non-positive score(s) are
not shown.

```{r distribution, fig.width=10, fig.height=4, eval=has_scores}
dist_data <- score_data %>%
  select(log_score, log_score_imputed) %>%
  pivot_longer(everything(), names_to = "stage", values_to = "value") %>%
  mutate(stage = factor(
    stage,
    levels = c("log_score", "log_score_imputed"),
    labels = c("Before imputation", "After imputation")
  ))

bound_line <- data.frame(
  stage = factor("Before imputation", levels = levels(dist_data$stage)),
  xintercept = lower_bound
)

ggplot(dist_data, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(data = bound_line, aes(xintercept = xintercept), color = "red") +
  facet_wrap(~stage) +
  labs(
    title = "Distribution of perturbation scores across all targets",
    x = "log10(perturbation score)",
    y = "Counts per bin"
  ) +
  theme_classic(base_size = 12)
```

```{r distribution-empty, eval=!has_scores, results="asis"}
cat("No positive perturbation score is available, so no distribution is shown.")
```

## Perturbation score heatmap

Scores are log10-transformed and quantile-cut as above; non-positive values are
shown as missing (gray).

```{r heatmap, fig.width=10, fig.height=max(4, 0.4 * n_cell_types)}
ggplot(score_data, aes(x = target, y = cell_type, fill = log_score_imputed)) +
  geom_tile(color = "white", linewidth = 0.6) +
  scale_fill_viridis_c(name = "log10(perb_score)", na.value = "gray80") +
  labs(x = "Target", y = "Cell type") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  { if (n_bindings > 1) facet_wrap(~binding) }
```

## Cell identity overview

The identities scored above, as they sit in the embedding. This is the cell set
after downsampling, so the group sizes are the ones the networks were actually
inferred from rather than those of the full object.

```{r umap, eval=has_umap, out.width="100%", fig.align="center"}
knitr::include_graphics(umap_path)
```

```{r umap-missing, eval=!has_umap, results="asis"}
cat("No UMAP figure was supplied with this report.")
```
