---
title: "NetPerturb Report"
subtitle: "Perturbation scores across targets and cell types"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    number-sections: true
    page-layout: full
    theme: flatly
params:
  perbscore_file: "perbscore_all_targets.txt"
  umap_file: "umap.png"
  network: ""
  score_quantile: 0.75
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(scales)

perb_score <- read.delim(params$perbscore_file, header = TRUE, sep = "\t")

n_cell_types <- n_distinct(perb_score$cell_type)
n_targets    <- n_distinct(perb_score$target)
n_bindings   <- n_distinct(perb_score$binding)

# Pool every score from every target on a log10 scale; non-positive scores
# cannot be logged and are carried as NA.
score_data <- perb_score %>%
  mutate(log_score = ifelse(perb_score > 0, log10(perb_score), NA_real_))

has_scores <- any(is.finite(score_data$log_score))

# Pooled scores are bimodal: a low mode of near-zero scores from genes the
# network barely leans on, and a high mode that carries the signal. Imputing
# the low mode up to a bound only stacked it against that bound and left it in
# the picture, so it is cut instead. Everything below the score_quantile
# quantile of the pooled log10 scores is dropped and the high mode is plotted
# on its own; the default 0.75 cuts at q3.
score_quantile <- suppressWarnings(as.numeric(params$score_quantile))
if (!isTRUE(is.finite(score_quantile)) || score_quantile < 0 || score_quantile >= 1) {
  score_quantile <- 0.75
}

score_cut <- if (has_scores) {
  unname(quantile(score_data$log_score, score_quantile, na.rm = TRUE))
} else {
  NA_real_
}

score_data <- score_data %>%
  mutate(log_score_kept = ifelse(
    is.finite(log_score) & log_score >= score_cut, log_score, NA_real_
  ))

n_scored  <- sum(is.finite(score_data$log_score))
n_kept    <- sum(is.finite(score_data$log_score_kept))
n_below   <- n_scored - n_kept
n_dropped <- sum(!is.finite(score_data$log_score))

# The network inference method the run was launched with (--network). The
# report also has to render when it is not supplied, e.g. when the qmd is
# knitted by hand against a table someone already has.
network_key   <- if (is.null(params$network)) "" else trimws(as.character(params$network))
network_names <- c(
  genie3  = "GENIE3",
  sctnet  = "scTenifoldNet",
  scrank  = "scRank",
  hdwgcna = "hdWGCNA"
)
network_label <- unname(network_names[network_key])
if (!nzchar(network_key)) {
  network_label <- "an unreported method"
} else if (is.na(network_label)) {
  network_label <- network_key
}
network_value <- if (nzchar(network_key)) network_key else "not supplied"

# The UMAP is a QC figure produced upstream by DOWNSAMPLE. A run started before
# that step existed, or one whose plot could not be drawn, still has to render.
umap_path <- params$umap_file
has_umap  <- !is.null(umap_path) && nzchar(umap_path) && file.exists(umap_path)
if (has_umap) umap_path <- normalizePath(umap_path)
```

## Overview

This report summarizes `r nrow(perb_score)` perturbation scores from
`r params$perbscore_file`, covering `r n_targets` target(s) across
`r n_cell_types` cell type(s), scored under `r n_bindings` binding
mode(s) (`r paste(sort(unique(perb_score$binding)), collapse = ", ")`).

The gene regulatory networks these scores were computed from were inferred with
**`r network_label`** (`--network`: `r network_value`). Every score below is
relative to that network type, so scores from a run using a different inference
method are not directly comparable.

## Perturbation score table

Use the column filters below to query by cell type, target, or binding.

```{r table}
perb_score %>%
  arrange(target, cell_type) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    options = list(pageLength = 15, order = list(list(1, "asc"))),
    caption = "All target x cell type perturbation scores."
  ) %>%
  formatSignif(columns = "perb_score", digits = 4)
```

## Perturbation score distribution

All `r n_scored` log10 scores from every target and cell type pooled into one
distribution. Pooled scores are typically bimodal: a low mode of near-zero
values from genes the network barely leans on, and a high mode carrying the
signal. Only the high mode is kept, cut at the `r score_quantile` quantile
(red line, `r if (has_scores) signif(score_cut, 4) else NA`). That retains
`r n_kept` score(s) and drops `r n_below` below the cut; a further `r n_dropped`
non-positive score(s) cannot be logged and are not shown either.

The left panel is every score, so the cut can be checked against where the two
modes actually separate — move it with `--score_quantile` if the red line does
not fall in the valley between them. The right panel is what the rest of the
report works with, on its own axis.

```{r distribution, fig.width=10, fig.height=4, eval=has_scores}
dist_data <- score_data %>%
  select(log_score, log_score_kept) %>%
  pivot_longer(everything(), names_to = "stage", values_to = "value") %>%
  filter(is.finite(value)) %>%
  mutate(stage = factor(
    stage,
    levels = c("log_score", "log_score_kept"),
    labels = c("All scores", "Retained high mode")
  ))

cut_line <- data.frame(
  stage = factor("All scores", levels = levels(dist_data$stage)),
  xintercept = score_cut
)

# A density curve is what makes the two modes readable, but it needs a few
# points per panel to mean anything.
show_density <- min(table(dist_data$stage)) >= 3

ggplot(dist_data, aes(x = value)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30, fill = "steelblue", color = "white"
  ) +
  { if (show_density) geom_density(color = "grey25", linewidth = 0.6) } +
  geom_vline(data = cut_line, aes(xintercept = xintercept), color = "red") +
  facet_wrap(~stage, scales = "free") +
  labs(
    title = "Distribution of perturbation scores across all targets",
    x = "log10(perturbation score)",
    y = "Density"
  ) +
  theme_classic(base_size = 12)
```

```{r distribution-empty, eval=!has_scores, results="asis"}
cat("No positive perturbation score is available, so no distribution is shown.")
```

## Perturbation score heatmap

Scores are log10-transformed and cut as above, so the colour scale spans the
high mode only. Tiles below the cut, and those whose score is non-positive, are
shown as missing (gray).

```{r heatmap, fig.width=10, fig.height=max(4, 0.4 * n_cell_types)}
ggplot(score_data, aes(x = target, y = cell_type, fill = log_score_kept)) +
  geom_tile(color = "white", linewidth = 0.6) +
  scale_fill_viridis_c(name = "log10(perb_score)", na.value = "gray80") +
  labs(x = "Target", y = "Cell type") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  { if (n_bindings > 1) facet_wrap(~binding) }
```

## Cell identity overview

The identities scored above, as they sit in the embedding. This is the cell set
after downsampling, so the group sizes are the ones the networks were actually
inferred from rather than those of the full object.

```{r umap, eval=has_umap, out.width="100%", fig.align="center"}
knitr::include_graphics(umap_path)
```

```{r umap-missing, eval=!has_umap, results="asis"}
cat("No UMAP figure was supplied with this report.")
```
