---
title: "NetPerturb Report"
subtitle: "Perturbation scores across targets and cell types"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    number-sections: true
    page-layout: full
    theme: flatly
params:
  perbscore_file: "perbscore_all_targets.txt"
  connections_file: "top_connections_all_targets.txt"
  umap_file: "umap.png"
  network: ""
  score_quantile: 0.75
  sctknk_file: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(scales)

# plotly draws the connection networks. A container built before that section
# existed still has to render the rest of the report, so its absence is a
# missing figure rather than an error.
has_plotly <- requireNamespace("plotly", quietly = TRUE)

perb_score <- read.delim(params$perbscore_file, header = TRUE, sep = "\t")

n_cell_types <- n_distinct(perb_score$cell_type)
n_targets    <- n_distinct(perb_score$target)
n_bindings   <- n_distinct(perb_score$binding)

# Pool every score from every target on a log10 scale; non-positive scores
# cannot be logged and are carried as NA.
score_data <- perb_score %>%
  mutate(log_score = ifelse(perb_score > 0, log10(perb_score), NA_real_))

has_scores <- any(is.finite(score_data$log_score))

# Pooled scores are bimodal: a low mode of near-zero scores from genes the
# network barely leans on, and a high mode that carries the signal. Imputing
# the low mode up to a bound only stacked it against that bound and left it in
# the picture, so it is cut instead. Everything below the score_quantile
# quantile of the pooled log10 scores is dropped and the high mode is plotted
# on its own; the default 0.75 cuts at q3.
score_quantile <- suppressWarnings(as.numeric(params$score_quantile))
if (!isTRUE(is.finite(score_quantile)) || score_quantile < 0 || score_quantile >= 1) {
  score_quantile <- 0.75
}

score_cut <- if (has_scores) {
  unname(quantile(score_data$log_score, score_quantile, na.rm = TRUE))
} else {
  NA_real_
}

score_data <- score_data %>%
  mutate(log_score_kept = ifelse(
    is.finite(log_score) & log_score >= score_cut, log_score, NA_real_
  ))

n_scored  <- sum(is.finite(score_data$log_score))
n_kept    <- sum(is.finite(score_data$log_score_kept))
n_below   <- n_scored - n_kept
n_dropped <- sum(!is.finite(score_data$log_score))

# The strongest scores in the run. "Overall" is the whole pooled set, the same
# population the retention cut is taken from, so this threshold does not move
# with score_quantile. Change top_quantile alone to mark a different fraction.
top_quantile <- 0.975
top_label    <- paste0("Top ", signif((1 - top_quantile) * 100, 3), "%")

top_cut <- if (has_scores) {
  unname(quantile(score_data$log_score, top_quantile, na.rm = TRUE))
} else {
  NA_real_
}

score_data <- score_data %>%
  mutate(is_top = is.finite(log_score) & log_score >= top_cut)

n_top <- sum(score_data$is_top)

# The network inference method the run was launched with (--network). The
# report also has to render when it is not supplied, e.g. when the qmd is
# knitted by hand against a table someone already has.
network_key   <- if (is.null(params$network)) "" else trimws(as.character(params$network))
network_names <- c(
  genie3  = "GENIE3",
  scrank  = "scRank",
  hdwgcna = "hdWGCNA"
)
network_label <- unname(network_names[network_key])
if (!nzchar(network_key)) {
  network_label <- "an unreported method"
} else if (is.na(network_label)) {
  network_label <- network_key
}
network_value <- if (nzchar(network_key)) network_key else "not supplied"

# The strongest edges RANK_SCORE pulled off each cell type network, one row per
# cell type x target gene x partner. A run started before RANK_SCORE emitted
# them has no such file.
connections_path <- params$connections_file
connections <- if (!is.null(connections_path) && nzchar(connections_path) &&
                   file.exists(connections_path)) {
  read.delim(connections_path, header = TRUE, sep = "\t")
} else {
  data.frame()
}

has_connections <- nrow(connections) > 0

if (has_connections) {
  connections <- connections %>%
    mutate(
      abs_weight = abs(weight),
      direction  = ifelse(weight >= 0, "Positive", "Negative")
    ) %>%
    arrange(target, cell_type, target_gene, rank)

  n_conn_targets    <- n_distinct(connections$target)
  n_conn_cell_types <- n_distinct(connections$cell_type)
  n_conn_views      <- nrow(distinct(connections, target, cell_type))
  max_kept          <- max(connections$rank)
} else {
  n_conn_targets    <- 0
  n_conn_cell_types <- 0
  n_conn_views      <- 0
  max_kept          <- 0
}

# The UMAP is a QC figure produced upstream by DOWNSAMPLE. A run started before
# that step existed, or one whose plot could not be drawn, still has to render.
umap_path <- params$umap_file
has_umap  <- !is.null(umap_path) && nzchar(umap_path) && file.exists(umap_path)
if (has_umap) umap_path <- normalizePath(umap_path)

# scTenifoldKnk's differentially-regulated (DR) gene table, from a --sctknk
# run. A header-only sentinel file (no rows) stands in when it was not
# requested for this run at all, same shape as a real run that just found
# nothing, so both render the same "nothing to show" message.
sctknk_path <- params$sctknk_file
sctknk_dr <- if (!is.null(sctknk_path) && nzchar(sctknk_path) &&
                  file.exists(sctknk_path)) {
  read.delim(sctknk_path, header = TRUE, sep = "\t")
} else {
  data.frame()
}

has_sctknk <- nrow(sctknk_dr) > 0

if (has_sctknk) {
  n_sctknk_targets    <- n_distinct(sctknk_dr$target)
  n_sctknk_cell_types <- n_distinct(sctknk_dr$cell_type)
  n_sctknk_sig        <- sum(sctknk_dr$p.adj < 0.05, na.rm = TRUE)
}
```

## Overview

This report summarizes `r nrow(perb_score)` perturbation scores from
`r params$perbscore_file`, covering `r n_targets` target(s) across
`r n_cell_types` cell type(s), scored under `r n_bindings` binding
mode(s) (`r paste(sort(unique(perb_score$binding)), collapse = ", ")`).

The gene regulatory networks these scores were computed from were inferred with
**`r network_label`** (`--network`: `r network_value`). Every score below is
relative to that network type, so scores from a run using a different inference
method are not directly comparable.

## Perturbation score table

Use the column filters below to query by cell type, target, or binding.

```{r table}
perb_score %>%
  arrange(target, cell_type) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    options = list(pageLength = 15, order = list(list(1, "asc"))),
    caption = "All target x cell type perturbation scores."
  ) %>%
  formatSignif(columns = "perb_score", digits = 4)
```

## Perturbation score distribution

All `r n_scored` log10 scores from every target and cell type pooled into one
distribution. Pooled scores are typically bimodal: a low mode of near-zero
values from genes the network barely leans on, and a high mode carrying the
signal. Only the high mode is kept, cut at the `r score_quantile` quantile
(red line, `r if (has_scores) signif(score_cut, 4) else NA`). That retains
`r n_kept` score(s) and drops `r n_below` below the cut; a further `r n_dropped`
non-positive score(s) cannot be logged and are not shown either.

The left panel is every score, so the cut can be checked against where the two
modes actually separate — move it with `--score_quantile` if the red line does
not fall in the valley between them. The right panel is what the rest of the
report works with, on its own axis.

The green dashed line in both panels is the `r top_label` threshold
(`r if (has_scores) signif(top_cut, 4) else NA`), the `r n_top` strongest
score(s) in the run. Those are the scores the heatmap asterisks mark.

```{r distribution, fig.width=10, fig.height=4, eval=has_scores}
dist_data <- score_data %>%
  select(log_score, log_score_kept) %>%
  pivot_longer(everything(), names_to = "stage", values_to = "value") %>%
  filter(is.finite(value)) %>%
  mutate(stage = factor(
    stage,
    levels = c("log_score", "log_score_kept"),
    labels = c("All scores", "Retained high mode")
  ))

# The retention cut belongs only to the panel it acts on; the top threshold is
# drawn in both, so the strongest scores can be placed against the whole set
# and against the mode the report keeps.
stages <- levels(dist_data$stage)

ref_lines <- rbind(
  data.frame(
    stage = factor("All scores", levels = stages),
    xintercept = score_cut,
    line = "Quantile cut"
  ),
  data.frame(
    stage = factor(stages, levels = stages),
    xintercept = top_cut,
    line = top_label
  )
)

# Red and green are a poor pair for the commonest colour blindness, so the two
# lines differ in dash as well as hue and the legend names both.
ref_colours   <- setNames(c("#e34948", "#008300"), c("Quantile cut", top_label))
ref_linetypes <- setNames(c("solid", "dashed"), c("Quantile cut", top_label))

# A density curve is what makes the two modes readable, but it needs a few
# points per panel to mean anything.
show_density <- min(table(dist_data$stage)) >= 3

ggplot(dist_data, aes(x = value)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 20, fill = "steelblue", color = "white"
  ) +
  { if (show_density) geom_density(color = "grey25", linewidth = 0.6) } +
  geom_vline(
    data = ref_lines,
    aes(xintercept = xintercept, colour = line, linetype = line),
    linewidth = 0.7
  ) +
  scale_colour_manual(name = NULL, values = ref_colours) +
  scale_linetype_manual(name = NULL, values = ref_linetypes) +
  facet_wrap(~stage, scales = "free") +
  labs(
    title = "Distribution of perturbation scores across all targets",
    x = "log10(perturbation score)",
    y = "Density"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r distribution-empty, eval=!has_scores, results="asis"}
cat("No positive perturbation score is available, so no distribution is shown.")
```

## Perturbation score heatmap

Scores are log10-transformed and cut as above, so the colour scale spans the
high mode only. Tiles below the cut, and those whose score is non-positive, are
shown as missing (gray). An asterisk marks a score in the `r top_label` of every
score in the run — `r n_top` tile(s), at or above
`r if (has_scores) signif(top_cut, 4) else NA` on the log10 scale.

```{r heatmap, fig.width=10, fig.height=max(4, 0.4 * n_cell_types)}
ggplot(score_data, aes(x = target, y = cell_type, fill = log_score_kept)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(
    data = filter(score_data, is_top),
    aes(label = "*"), colour = "#0b0b0b", size = 6, vjust = 0.75
  ) +
  scale_fill_viridis_c(name = "log10(perb_score)", na.value = "gray80") +
  labs(x = "Target", y = "Cell type") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  { if (n_bindings > 1) facet_wrap(~binding) }
```

## Target connection networks

The strongest edges each target holds in each cell type network, taken from the
target's row of that network — the same neighbourhood `scRank::init_mod()` reads
when it modularises a subnetwork, but kept per cell type and capped at the
`r max_kept` strongest edges. Use the menu to switch between the
`r n_conn_views` target and cell type combinations.

The target sits at the centre and its partners around it. An edge's colour and
dash carry the sign of the weight, a solid blue edge being a positive one and a
dashed red edge a negative one; a partner's size follows the strength of its
edge, scaled within the combination on display, so sizes rank partners against
each other rather than across cell types. Hover any node or edge for the exact
weight; every value is also in the table below.

```{r network, eval=has_connections && has_plotly}
# Blue and red are reserved for the sign of the weight, the one thing on this
# figure carrying meaning through colour, so the nodes stay neutral.
edge_pos_colour <- "#2a78d6"
edge_neg_colour <- "#e34948"
node_fill       <- "#e8e7e3"
node_line       <- "#52514e"
target_fill     <- "#3f3e3b"
surface         <- "#fcfcfb"

# The subnetwork of one target is a star, so it needs no force layout: the
# target genes go in the middle and their partners onto a ring around them.
star_positions <- function(edges) {
  centre <- sort(unique(edges$target_gene))
  ring   <- setdiff(unique(edges$partner), centre)

  centre_pos <- if (length(centre) == 1) {
    data.frame(gene = centre, x = 0, y = 0, label_at = "bottom center")
  } else {
    angle <- (seq_along(centre) - 1) * 2 * pi / length(centre) + pi / 2
    data.frame(gene = centre, x = 0.22 * cos(angle), y = 0.22 * sin(angle),
               label_at = "bottom center")
  }

  ring_pos <- if (length(ring) > 0) {
    angle <- (seq_along(ring) - 1) * 2 * pi / length(ring) + pi / 2
    # A label pushed straight out from the centre clears both its own node and
    # the edge running into it, which "top center" on every node does not.
    label_at <- ifelse(
      cos(angle) > 0.5, "middle right",
      ifelse(cos(angle) < -0.5, "middle left",
             ifelse(sin(angle) > 0, "top center", "bottom center"))
    )
    data.frame(gene = ring, x = cos(angle), y = sin(angle), label_at = label_at)
  } else {
    data.frame(gene = character(), x = numeric(), y = numeric(),
               label_at = character())
  }

  rbind(centre_pos, ring_pos)
}

# Every segment of a plotly line trace shares one colour and width, so all the
# edges of one sign are drawn as a single trace with NA breaking them apart.
edge_path <- function(edges, pos) {
  if (nrow(edges) == 0) {
    return(list(x = NA_real_, y = NA_real_))
  }
  from <- pos[match(edges$target_gene, pos$gene), ]
  to   <- pos[match(edges$partner, pos$gene), ]
  list(
    x = as.vector(rbind(from$x, to$x, NA_real_)),
    y = as.vector(rbind(from$y, to$y, NA_real_))
  )
}

marker_size <- function(w, lo = 16, hi = 34) {
  if (length(w) == 0) return(numeric(0))
  span <- range(w)
  if (!is.finite(diff(span)) || diff(span) == 0) return(rep((lo + hi) / 2, length(w)))
  lo + (w - span[1]) / diff(span) * (hi - lo)
}

views <- connections %>%
  distinct(target, cell_type) %>%
  arrange(target, cell_type) %>%
  mutate(label = paste0(target, "  ·  ", cell_type))

fig <- plotly::plot_ly(width = NULL, height = 640)

# The dropdown hides and shows the real traces, so their legend entries would
# come and go with them. These two carry the legend instead and never move.
# They are parked far outside the fixed axis range rather than left empty: a
# trace with no coordinates loses its name, and with it its legend entry.
legend_x <- c(-99, -98)
legend_y <- c(-99, -99)

fig <- fig %>%
  plotly::add_trace(
    x = legend_x, y = legend_y, type = "scatter", mode = "lines",
    line = list(color = edge_pos_colour, width = 2.5),
    name = "Positive weight", hoverinfo = "skip", showlegend = TRUE
  ) %>%
  plotly::add_trace(
    x = legend_x, y = legend_y, type = "scatter", mode = "lines",
    line = list(color = edge_neg_colour, width = 2.5, dash = "dash"),
    name = "Negative weight", hoverinfo = "skip", showlegend = TRUE
  )

view_traces <- vector("list", nrow(views))
next_trace  <- 2L  # the two legend traces above

for (i in seq_len(nrow(views))) {
  edges <- connections %>%
    filter(target == views$target[i], cell_type == views$cell_type[i])

  pos <- star_positions(edges)
  visible <- i == 1

  positive <- edge_path(filter(edges, weight >= 0), pos)
  negative <- edge_path(filter(edges, weight < 0), pos)

  hover <- paste0(
    edges$target_gene, " — ", edges$partner,
    "<br>weight ", signif(edges$weight, 4),
    "<br>rank ", edges$rank
  )
  mid <- data.frame(
    x = (pos$x[match(edges$target_gene, pos$gene)] + pos$x[match(edges$partner, pos$gene)]) / 2,
    y = (pos$y[match(edges$target_gene, pos$gene)] + pos$y[match(edges$partner, pos$gene)]) / 2
  )

  # A partner reached by more than one target gene is drawn once, at its
  # strongest edge.
  partners <- edges %>%
    group_by(partner) %>%
    summarise(abs_weight = max(abs_weight), .groups = "drop") %>%
    left_join(pos, by = c(partner = "gene"))

  partner_hover <- edges %>%
    group_by(partner) %>%
    summarise(
      text = paste0(
        partner[1], "<br>",
        paste0(target_gene, " — ", signif(weight, 4), collapse = "<br>")
      ),
      .groups = "drop"
    )
  partners <- left_join(partners, partner_hover, by = "partner")

  centres <- pos[pos$gene %in% edges$target_gene, ]

  fig <- fig %>%
    plotly::add_trace(
      x = positive$x, y = positive$y, type = "scatter", mode = "lines",
      line = list(color = edge_pos_colour, width = 2.5),
      hoverinfo = "skip", showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = negative$x, y = negative$y, type = "scatter", mode = "lines",
      line = list(color = edge_neg_colour, width = 2.5, dash = "dash"),
      hoverinfo = "skip", showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = mid$x, y = mid$y, type = "scatter", mode = "markers",
      marker = list(size = 14, color = "rgba(0,0,0,0)"),
      text = hover, hoverinfo = "text", showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = partners$x, y = partners$y, type = "scatter", mode = "markers+text",
      marker = list(
        size = marker_size(partners$abs_weight),
        color = node_fill, opacity = 1,
        line = list(color = node_line, width = 1.5)
      ),
      text = partners$partner, textposition = partners$label_at,
      textfont = list(size = 11, color = "#0b0b0b"),
      hovertext = partners$text, hoverinfo = "text",
      showlegend = FALSE, visible = visible
    ) %>%
    plotly::add_trace(
      x = centres$x, y = centres$y, type = "scatter", mode = "markers+text",
      marker = list(
        size = 34, color = target_fill, opacity = 1,
        line = list(color = surface, width = 2)
      ),
      text = centres$gene, textposition = "bottom center",
      textfont = list(size = 13, color = "#0b0b0b"),
      hovertext = paste0(centres$gene, "<br>target"), hoverinfo = "text",
      showlegend = FALSE, visible = visible
    )

  view_traces[[i]] <- next_trace + seq_len(5)
  next_trace <- next_trace + 5L
}

n_traces <- next_trace

# The title is restated on every button: a layout update replaces the title
# whole, so leaving out its placement would recentre it on the first switch.
title_style <- function(text) {
  list(text = text, x = 0, xanchor = "left",
       font = list(size = 15, color = "#0b0b0b"))
}

buttons <- lapply(seq_len(nrow(views)), function(i) {
  visible <- rep(FALSE, n_traces)
  visible[1:2] <- TRUE
  visible[view_traces[[i]]] <- TRUE

  list(
    method = "update",
    args = list(
      list(visible = as.list(visible)),
      list(title = title_style(views$label[i]))
    ),
    label = views$label[i]
  )
})

fig %>%
  plotly::layout(
    title = title_style(views$label[1]),
    # No aspect lock: letting the ring stretch to the page width spreads the
    # partner labels apart instead of leaving the space either side empty.
    xaxis = list(visible = FALSE, range = c(-1.75, 1.75)),
    yaxis = list(visible = FALSE, range = c(-1.35, 1.35)),
    legend = list(orientation = "h", x = 0, y = -0.02,
                  font = list(color = "#52514e")),
    margin = list(l = 20, r = 20, t = 110, b = 40),
    paper_bgcolor = surface,
    plot_bgcolor = surface,
    hoverlabel = list(bgcolor = "#ffffff", bordercolor = "#d6d5d0"),
    updatemenus = list(list(
      type = "dropdown", direction = "down", active = 0,
      x = 1, xanchor = "right", y = 1.12, yanchor = "top",
      bgcolor = "#ffffff", bordercolor = "#d6d5d0",
      font = list(size = 12), buttons = buttons
    ))
  ) %>%
  plotly::config(displayModeBar = FALSE)
```

```{r network-missing, eval=!(has_connections && has_plotly), results="asis"}
if (!has_connections) {
  cat("No top connection table was supplied with this report, so no network is shown.")
} else {
  cat("plotly is not installed in this container, so the connection networks",
      "could not be drawn. The table below still lists every connection.")
}
```

## Top connections table

Every edge RANK_SCORE kept: up to `r max_kept` per target gene per cell type,
across `r n_conn_targets` target(s) and `r n_conn_cell_types` cell type(s).
`weight` is the edge weight in that cell type's network, signed, and `rank`
orders the edges of one target gene in one cell type by absolute weight.

```{r connections-table, eval=has_connections}
connections %>%
  select(cell_type, target, target_gene, partner, weight, rank) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    options = list(pageLength = 15, order = list(list(5, "asc"))),
    caption = "Top connections per target gene per cell type."
  ) %>%
  formatSignif(columns = "weight", digits = 4)
```

```{r connections-table-empty, eval=!has_connections, results="asis"}
cat("No top connection table was supplied with this report.")
```

## scTenifoldKnk differentially-regulated genes

A separate, target-specific analysis: for each cell type x target pair,
scTenifoldKnk builds the wild-type network, knocks the target out of it
directly (zeroing its outgoing edges), and compares the two networks by
manifold alignment. `distance` is how far a gene moves between the two;
`p.value`/`p.adj` (Benjamini-Hochberg) test that distance against the
chi-square null. This does not produce a `perb_score` and is not folded into
any of the tables or figures above -- it is a separate way of asking the same
question. Currently `r if (has_sctknk) n_sctknk_sig else 0` gene(s) clear
FDR < 0.05, across `r if (has_sctknk) n_sctknk_targets else 0` target(s) and
`r if (has_sctknk) n_sctknk_cell_types else 0` cell type(s).

```{r sctknk-table, eval=has_sctknk}
sctknk_dr %>%
  arrange(target, cell_type, p.value) %>%
  select(cell_type, target, gene, distance, p.value, p.adj) %>%
  datatable(
    filter = "top",
    rownames = FALSE,
    options = list(pageLength = 15, order = list(list(4, "asc"))),
    caption = "scTenifoldKnk differentially-regulated genes per cell type x target."
  ) %>%
  formatSignif(columns = c("distance", "p.value", "p.adj"), digits = 4)
```

```{r sctknk-table-empty, eval=!has_sctknk, results="asis"}
cat("scTenifoldKnk was not run for this session, so no differentially-regulated",
    "gene table is shown. Pass --sctknk to generate one.")
```

## Cell identity overview

The identities scored above, as they sit in the embedding. This is the cell set
after downsampling, so the group sizes are the ones the networks were actually
inferred from rather than those of the full object.

```{r umap, eval=has_umap, out.width="100%", fig.align="center"}
knitr::include_graphics(umap_path)
```

```{r umap-missing, eval=!has_umap, results="asis"}
cat("No UMAP figure was supplied with this report.")
```
